---
title: "Getting started with dynR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dynR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(dynR)
```

## What is dynamic functional connectivity?

Resting-state fMRI measures the BOLD signal across brain regions over time.
**Static** functional connectivity (FC) summarises this as a single
correlation matrix — the average co-activation pattern across the entire scan.
**Dynamic** functional connectivity (dynFC) asks a different question: does
that pattern change over time, and if so, how?

`dynR` computes dynFC representations from preprocessed BOLD timeseries. It
implements two complementary families of methods:

| Family | Key functions | What it captures |
|---|---|---|
| Phase-based | `hilbert_phases()`, `dyn_phase_lock()`, `kuramoto()` | Instantaneous synchrony and phase relationships |
| Correlation-based | `corr_slide()`, `cofluct()`, `corr_corr()` | Windowed or edge-level co-activation |

The outputs of `dynR` feed directly into
[`stateR`](https://github.com/CoDe-Neuro/stateR) for brain state
quantification (fractional occupancy, dwell time, Markov transitions).

---

## Package data

`dynR` ships with a resting-state BOLD timeseries from the
[edge-ts](https://github.com/brain-networks/edge-ts) repository, parcellated
into 200 regions:

```{r load-data}
data(ts)   # [200 x 600] BOLD timeseries
data(fc)   # [200 x 200] static FC matrix (ground truth)

dim(ts)
dim(fc)
```

200 parcels, 600 timepoints. We will use this throughout the vignettes.

---

## Pipeline overview

A typical dynR workflow looks like this:

```
BOLD timeseries [N × Tmax]
        │
        ├─── bandpass_filter()       ← optional preprocessing
        │
        ├─── Phase-based ────────────────────────────────────────
        │     hilbert_phases()       → instantaneous phases [N × Tmax]
        │     dyn_phase_lock()       → phase-locking matrices + LEiDA vectors
        │     kuramoto()             → synchrony, metastability, entropy
        │
        └─── Correlation-based ──────────────────────────────────
              corr_slide()          → sliding-window FC [N × N × windows]
              cofluct()             → edge time series + RSS
              corr_corr()           → correlation-of-correlations [Tmax × Tmax]
                    │
                    ▼
              stateR (K-means → brain states)
```

---

## Quick example: phase-based pipeline

```{r phase-pipeline}
# 1. Extract instantaneous phases via the Hilbert transform
phases <- hilbert_phases(ts)
dim(phases)  # same as ts: [200 x 600]

# 2. Dynamic phase-locking matrices + leading eigenvectors (LEiDA)
dpl <- dyn_phase_lock(phases)
dim(dpl$sync_conn)  # [200 x 200 x 580]  (trimmed 10 timepoints each end)
dim(dpl$leida)      # [580 x 200]         leading eigenvectors

# 3. Kuramoto order parameter
kop <- kuramoto(phases)
cat("Metastability:", round(kop$metastability, 4), "\n")
cat("Entropy:      ", round(kop$entropy, 4), "\n")
```

The `leida` matrix — one leading eigenvector per timepoint — is ready to
pass to K-means for brain state discovery.

---

## Quick example: sliding-window pipeline

```{r sw-pipeline}
# Sliding-window correlation: 30-timepoint windows, step of 5
sw <- corr_slide(ts, window = 30, step = 5)
cat("Number of windows:", dim(sw$corr_mats)[3], "\n")

# Edge-centric cofluctuations
ec <- cofluct(ts)
cat("Edge time series shape:", nrow(ec$edge_ts), "x", ncol(ec$edge_ts), "\n")
cat("RSS range: [", round(min(ec$rss), 2), ",", round(max(ec$rss), 2), "]\n")
```

---

## Validating against static FC

A useful sanity check: a single window spanning the full timeseries should
reproduce the static FC matrix exactly.

```{r sanity-check}
full_window <- corr_slide(ts, window = ncol(ts))
max_diff    <- max(abs(full_window$corr_mats[, , 1] - fc))
cat("Max deviation from static FC:", formatC(max_diff, format = "e"), "\n")
```

---

## What comes next

- `vignette("phase-based-fc")` — Hilbert transform, LEiDA, and the Kuramoto order parameter in depth
- `vignette("sliding-window-fc")` — sliding-window correlation and edge-centric cofluctuations
- Passing `dpl$leida` to `kmeans()` and `stateR` for brain state quantification
