---
title: "Phase-based dynamic FC: Hilbert transform, LEiDA, and Kuramoto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phase-based dynamic FC: Hilbert transform, LEiDA, and Kuramoto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(dynR)
data(ts)
```

## Overview

Phase-based dynFC methods characterise functional connectivity through the
**instantaneous phase** of the BOLD signal rather than its amplitude. The
analytic signal (Hilbert transform) decomposes each parcel's timeseries into
a phase and an envelope, and the phase relationships across parcels carry
information about transient synchronisation states.

`dynR` implements three phase-based methods:

| Function | Output | Method |
|---|---|---|
| `hilbert_phases()` | Instantaneous phases \[N × Tmax\] | Hilbert transform |
| `dyn_phase_lock()` | Phase-locking matrices + LEiDA vectors | dPL / LEiDA |
| `kuramoto()` | Synchrony, metastability, entropy | Kuramoto order parameter |

---

## Step 1: Bandpass filtering

Before extracting phases, it is standard practice to bandpass filter the BOLD
signal to the frequency range of interest. For resting-state fMRI with a TR
of 2 s, the typical band is 0.01–0.1 Hz.

```{r filter}
ts_filt <- apply(ts, 1, bandpass_filter,
                 flp = 0.01, fhi = 0.1, delt = 2, order = 2)
ts_filt <- t(ts_filt)   # restore [N x Tmax] orientation
dim(ts_filt)
```

The `apply` call filters each parcel (row) independently; `t()` restores the
\[N × Tmax\] convention expected by downstream functions.

---

## Step 2: Instantaneous phases via the Hilbert transform

`hilbert_phases()` computes the analytic signal of each (demeaned) parcel
timeseries and extracts the instantaneous phase angle.

```{r hilbert}
phases <- hilbert_phases(ts_filt)
dim(phases)   # [200 x 600]
range(phases) # in [-pi, pi]
```

Each row is a parcel; each column is a timepoint. Values are in radians.

---

## Step 3: Dynamic phase-locking matrix (dPL) and LEiDA

For each timepoint *t*, `dyn_phase_lock()` constructs an *N* × *N*
**phase-locking matrix** whose entry *(i, j)* is:

$$\text{dPL}_{ij}(t) = \cos(\phi_i(t) - \phi_j(t))$$

where *φ* denotes instantaneous phase. A value near 1 indicates the two
parcels are phase-locked; a value near −1 indicates anti-phase synchrony.

The **leading eigenvector** (LEiDA) of each dPL matrix captures the dominant
large-scale synchronisation pattern at that timepoint (Cabral et al., 2017).
Ten timepoints are trimmed from each end to avoid Hilbert transform edge
effects.

```{r dpl}
dpl <- dyn_phase_lock(phases)

# Phase-locking array: [200 x 200 x 580]
dim(dpl$sync_conn)

# LEiDA matrix: [580 x 200]  — one eigenvector per (trimmed) timepoint
dim(dpl$leida)
```

### Inspecting a single phase-locking matrix

```{r dpl-slice}
# Symmetry and diagonal check on the first timepoint
t1 <- dpl$sync_conn[, , 1]
cat("Symmetric:", isTRUE(all.equal(t1, t(t1))), "\n")
cat("Diagonal:", round(unique(diag(t1)), 6), "\n")   # cos(0) = 1
```

---

## Step 4: Brain state discovery with K-means

The LEiDA vectors are the natural input to K-means clustering. Each cluster
centre represents a recurring whole-brain phase-locking pattern — a
**brain state**.

```{r leida-kmeans}
set.seed(42)
K  <- 5
km <- kmeans(dpl$leida, centers = K, nstart = 50, iter.max = 500)

table(km$cluster)
```

The resulting state sequence (`km$cluster`) can be passed directly to
`stateR` for fractional occupancy, dwell time, and Markov transition
analysis.

---

## Step 5: Kuramoto order parameter

The **Kuramoto order parameter** *R(t)* measures the instantaneous global
synchrony across all parcels:

$$R(t) = \frac{1}{N} \left| \sum_{j=1}^{N} e^{i\phi_j(t)} \right|$$

*R* = 1 means all parcels are perfectly phase-locked; *R* = 0 means phases
are uniformly distributed. `kuramoto()` returns the full synchrony time
series along with two summary scalars:

- **Metastability** — standard deviation of *R(t)*; reflects how much the
  global synchronisation level fluctuates over time.
- **Entropy** — Shannon entropy of the discretised synchrony distribution;
  captures the diversity of synchrony states visited.

```{r kuramoto}
kop <- kuramoto(phases, base = 2, n_bits = 8)

cat("Metastability:", round(kop$metastability, 4), "\n")
cat("Entropy (bits):", round(kop$entropy, 4), "\n")
cat("Synchrony range: [",
    round(min(kop$synchrony), 3), ",",
    round(max(kop$synchrony), 3), "]\n")
```

### Visualising the synchrony time series

```{r kuramoto-plot, fig.width = 7, fig.height = 3}
plot(kop$synchrony, type = "l", col = "#1B6799",
     xlab = "Timepoint", ylab = "R(t)",
     main = "Kuramoto order parameter",
     ylim = c(0, 1), lwd = 1.2)
abline(h = mean(kop$synchrony), col = "#FC544A", lty = 2, lwd = 1.5)
legend("topright", legend = "Mean R(t)", col = "#FC544A",
       lty = 2, lwd = 1.5, bty = "n")
```

---

## Summary

```{r summary}
cat("Parcels:              ", nrow(ts_filt), "\n")
cat("Timepoints (raw):     ", ncol(ts_filt), "\n")
cat("Timepoints (trimmed): ", nrow(dpl$leida), "\n")
cat("Brain states (K):     ", K, "\n")
cat("Metastability:        ", round(kop$metastability, 4), "\n")
```

---

## References

Cabral, J. et al. (2017). Cognitive performance in healthy older adults
relates to spontaneous switching between states of functional connectivity
during rest. *Scientific Reports*, 7(1), 5135.
https://doi.org/10.1038/s41598-017-05425-7

Lord, L.-D. et al. (2019). Dynamical exploration of the repertoire of brain
networks at rest is modulated by psilocybin. *NeuroImage*, 199, 127–142.
https://doi.org/10.1016/j.neuroimage.2019.05.060
