---
title: "Correlation-based dynamic FC: sliding windows and edge cofluctuations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlation-based dynamic FC: sliding windows and edge cofluctuations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(dynR)
data(ts)
data(fc)
```

## Overview

Correlation-based dynFC methods characterise how functional connectivity
evolves over time by either **windowing** the timeseries into short segments
and computing Pearson correlations within each window, or by analysing the
**instantaneous co-activation** of parcel pairs as a continuous edge time
series.

`dynR` implements three correlation-based functions:

| Function | Output | Method |
|---|---|---|
| `corr_slide()` | FC matrices \[N × N × windows\] | Sliding-window Pearson correlation |
| `cofluct()` | Edge time series \[n\_edges × Tmax\] + RSS | Edge-centric cofluctuation |
| `corr_corr()` | Correlation-of-correlations \[Tmax × Tmax\] | Hansen et al. (2015) |

---

## Sliding-window correlation

`corr_slide()` divides the timeseries into overlapping or non-overlapping
windows and computes the Pearson FC matrix within each one. The choice of
window length is a key parameter: too short and the estimate is noisy; too
long and rapid transitions are blurred.

```{r corr-slide}
# 30-timepoint windows, step of 5 (overlapping)
sw <- corr_slide(ts, window = 30, step = 5)

cat("Window size: 30 timepoints\n")
cat("Step:        5 timepoints\n")
cat("N windows:  ", dim(sw$corr_mats)[3], "\n")
```

### Validating against static FC

When `window = ncol(ts)` the single window is the full timeseries, so the
result must equal the static FC matrix.

```{r sw-validation}
full <- corr_slide(ts, window = ncol(ts))
max_diff <- max(abs(full$corr_mats[, , 1] - fc))
cat("Max deviation from static FC:", formatC(max_diff, format = "e"), "\n")
```

### FC variability across windows

A useful summary is how much the inter-parcel correlations vary across
windows. High variability means the connectivity structure is genuinely
non-stationary.

```{r sw-variability}
# Upper triangle indices (excluding diagonal)
n_parcels <- dim(sw$corr_mats)[1]
ut <- which(upper.tri(diag(n_parcels)), arr.ind = TRUE)

# Extract upper triangle for each window → [n_edges x n_windows] matrix
edge_by_window <- apply(sw$corr_mats, 3, function(m) m[ut])

# Mean standard deviation across edges
mean_sd <- mean(apply(edge_by_window, 1, sd))
cat("Mean edge SD across windows:", round(mean_sd, 4), "\n")
```

---

## Edge-centric cofluctuations

The edge-centric framework (Esfahlani et al., 2020; Faskowitz et al., 2020)
does not window the data at all. Instead, it z-scores each parcel's
timeseries and computes the element-wise product for every unique parcel pair:

$$\text{ets}_{ij}(t) = z_i(t) \cdot z_j(t)$$

The result is an **edge time series** — a continuous, frame-by-frame estimate
of co-activation for every pair.

```{r cofluct}
ec <- cofluct(ts)

n_edges <- nrow(ec$edge_ts)
cat("Edges (N*(N-1)/2):", n_edges, "\n")   # 200*199/2 = 19900
cat("Timepoints:       ", ncol(ec$edge_ts), "\n")
```

### Root-sum-square (RSS) cofluctuation

The RSS vector summarises the total cofluctuation amplitude at each
timepoint:

$$\text{RSS}(t) = \sqrt{\sum_{(i,j)} \text{ets}_{ij}(t)^2}$$

High-RSS frames are moments of unusually strong co-activation across the
brain and have been shown to drive the structure of the static FC matrix
(Esfahlani et al., 2020).

```{r rss-plot, fig.width = 7, fig.height = 3}
plot(ec$rss, type = "l", col = "#1B6799",
     xlab = "Timepoint", ylab = "RSS",
     main = "Root-sum-square cofluctuation",
     lwd = 1.2)
abline(h = mean(ec$rss) + 2 * sd(ec$rss),
       col = "#FC544A", lty = 2, lwd = 1.5)
legend("topright", legend = "Mean + 2 SD",
       col = "#FC544A", lty = 2, lwd = 1.5, bty = "n")
```

### High-amplitude events

Frames with RSS above mean + 2 SD are candidate **high-amplitude cofluctuation
events** — the moments that most strongly shape static FC.

```{r high-amp}
threshold  <- mean(ec$rss) + 2 * sd(ec$rss)
high_frames <- which(ec$rss > threshold)
cat("High-amplitude frames (RSS > mean + 2SD):", length(high_frames), "\n")
cat("Proportion of scan:", round(length(high_frames) / ncol(ts), 3), "\n")
```

---

## Correlation of correlations

`corr_corr()` computes the full \[Tmax × Tmax\] matrix of pairwise
correlations between edge time series (Hansen et al., 2015). Entry *(t1, t2)*
reflects how similar the co-activation patterns were at timepoints *t1* and
*t2* — a non-windowed measure of FC recurrence.

```{r corr-corr}
cc <- corr_corr(ts)
dim(cc)   # [600 x 600]
```

The matrix is symmetric by construction and has ones on the diagonal.

```{r cc-check}
cat("Symmetric:   ", isTRUE(all.equal(cc, t(cc))), "\n")
cat("Diagonal == 1:", all(abs(diag(cc) - 1) < 1e-10), "\n")
```

---

## Preparing for brain state analysis

Both sliding-window FC and the correlation-of-correlations matrix can be used
for K-means brain state discovery. The standard approach with sliding-window
FC is to vectorise the upper triangle of each window's FC matrix and cluster
the resulting feature vectors:

```{r sw-clustering}
# Vectorise upper triangle for each window
ut       <- which(upper.tri(diag(n_parcels)), arr.ind = TRUE)
features <- t(apply(sw$corr_mats, 3, function(m) m[ut]))
# features: [n_windows x n_edges]

set.seed(42)
K  <- 5
km <- kmeans(features, centers = K, nstart = 50, iter.max = 500)
table(km$cluster)
```

The state sequence can then be passed to `stateR` for fractional occupancy,
dwell time, and Markov transition analysis.

---

## References

Hansen, E. C. A. et al. (2015). Functional connectivity dynamics: Modeling
the switching behavior of the resting state. *NeuroImage*, 105, 525–535.
https://doi.org/10.1016/j.neuroimage.2014.11.001

Esfahlani, F. Z. et al. (2020). High-amplitude cofluctuations in cortical
activity drive functional connectivity. *PNAS*, 117(45), 28393–28401.
https://doi.org/10.1073/pnas.2005531117

Faskowitz, J. et al. (2020). Edge-centric functional network representations
of human cerebral cortex reveal overlapping system-level architecture.
*Nature Neuroscience*, 23(12), 1644–1654.
https://doi.org/10.1038/s41593-020-00719-y
